\FOO: A framework for efficient source code generation\\ of detection algorithms for the Internet of Things

### 2015-04-16 paper-euspn15.tex
cleaned up the paper up to 9 pages (1 to go)

\title{
\NAME: A framework for Efficient Intrusion Detection in the Internet of Things
}

\author[a]{Christophe Van Ginneken}
\author[a]{Jef Maerien}
\author[a]{Christophe Huygens}
\author[a]{\\Wouter Joosen}
\author[a]{Danny Hughes}

\address[a]{
iMinds-DistriNet, KU Leuven, 3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be
}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Manual fusion of the algorithms reduces resource
consumption by optimizing resource sharing, yet, proves to be time-consuming,
repetitive and error-prone. To address this problem we propose \NAME, a
framework for the development of efficient intrusion detection systems. It
consists of a domain specific language and a code generator. The language
allows formally describing the intent of intrusion detection algorithms and
supports the generator in organizing the source code. A side-by-side comparison
shows that \NAME-generated code reduces message passing overhead, execution
time and memory footprint in comparison to sequential calls into individual
implementations of the algorithms.

@@

IoT, Domain Specific Languages, Code Generation, Source Code Optimization,
Intrusion Detection

@@

% context

% wired networks: fw and central ids

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. Still,
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

% IoT + threat

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health, enhanced
learning, etc. \cite{atzori2010internet}. With this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords \& encryption
keys). Implementation of security measures by itself is hindered by the fact
that IoT devices typically have limited batteries, processing power and memory.

% local ids -> localize -> limited resources

In wireless networks of resource-constrained devices, which make up a
significant part of the IoT, it is not possible for a single point in the
network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
implement its own lines of defense. However making the IDS a local service on
each device requires local resources, which are a scarce commodity for IoT
devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the physical layer, through the
access and routing layers, up to the application layer. Each layer presents
different opportunities to manipulate data, eavesdrop or perform a form of
denial of service. For each of these attacks an algorithm needs to look for
patterns or anomalies. An IDS for IoT devices therefore requires a large number
of algorithms to achieve adequate coverage.

% gap analysis

Current developments in ID on resource-constrained devices focus on programming
frameworks that structure the implementation of algorithms \cite{valero2012di}
and offer the required basic functional components to implement them
\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks is that
they do not offer a way to optimize the usage of resources. A solution that
supports the integration of multiple algorithms on IoT devices needs to avoid
accumulating their impact on the available resources. A second omission is
support for systematic reuse to create different configurations. Addressing
this in a transparent and automated way offers the potential to significantly
reduce the effort of developing an IDS.

@@

% 1. classification of ID and design pattern for ID algorithms

The first contribution of this paper consists in the formulation of a design
pattern for ID algorithms, based on properties identified from a classification.

% 2. framework: language + code generator

The second contribution is a framework called \NAME. It enables the formal
description of ID algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \NAME addresses the heterogeneous nature of both IoT and ID algorithms.

% 3. FCF to optimise

The third contribution introduces functional code fusion (FCF) as a source code
generation paradigm to address hard to combine ID algorithms. FCF identifies
common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
describes the inherent problems in the combination of ID algorithms. Section
\ref{classification} analyses and classifies ID algorithms. Section
\ref{pattern} looks for patterns in the implementation of these classes.
Section \ref{design} describes the design we applied to construct \NAME, its
domain specific language (DSL) and code generator. Section \ref{evaluation}
discusses and evaluates an implementation. Section \ref{related} explores
related work in the field of DSLs and code generation. Finally, section
\ref{conclusion} summarizes our findings, draws conclusions and identifies
topics for future work.

\section{Inherent Problems}
\label{problems}

The combination of ID algorithms tends to result in suboptimal code. Besides
these code-level problems, operational problems also hinder the implementation.

Two issues arise from the way ID algorithms are implemented and deal with loop
constructs and common data. A third issue is the use of the network.

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can cause overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns.

  \item[Common data] cross-cuts the different algorithms, that have to store
  information about devices within its own scope. This results in a lot of
  duplicated information in memory.

  \item[The network] allows algorithms to operate on a distributed scale.
  However, scattered calls to a message sending function cause multiple
  accesses to the wireless radio, keeping it active and energy consuming more
  than strictly needed.

@@

Two operational aspects are orthogonal to these problems and amplify them:

@@

  \item[Reimplementing] the algorithms manually to eliminate these issues is
  not a one-time cost. A release of a new version of an algorithm, or simply a
  reselection of algorithms, due to a changing risk analysis, can trigger it.
  Also, changing, or simply implementing the algorithms from scratch, holds the
  risk of making mistakes.

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This partially explains why
  there are few implementations readily available to developers: there is no
  single platform to target for researchers.

@@

\section{A Classification of ID Algorithms}
\label{classification}

Two aspects currently missing from ID frameworks are optimization of resources
and support for systematic reuse. This section uses a classification to define
the scope of \NAME, detailing which types of algorithms are good candidates for
the proposed approach.

Research into ID in wireless networks typically focuses on two major topics:
the devices as single entities and the network as a group of such devices. Both
are important: a group cannot make a decision without members that detect
malicious behavior and a group-based decision is often needed because devices
in the network can miss certain events that could reveal intrusions. In a
wireless network, where not all participants are within each others' radio
range, wireless devices typically only have a partial view on the entire
network. This prohibits them from observing potentially crucial events in
distant parts of the network. The fact that resource-constrained devices
typically apply duty cycling to lower their energy consumption, further
increases this problem, introducing periods of time when they cannot observe
local events.

This leads to two complementary classes of ID activities: local intrusion
detection and cooperative decision making.

\subsection{Local Intrusion Detection}
\label{detection}

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: 1) anomaly
detection, 2) signature or misuse detection, and 3) specification-based
detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
and add the notion of hybrid intrusion detection systems and cross layer
intrusion detection systems as recent advances in research. Although these
additional categories seem promising, the authors admit that the impact on
resource-constrained devices will be too high. We therefore focus on the three
categories agreed upon by most sources.

  \item[Anomaly detection] considers a baseline profile and evaluates all
  activity according to this profile. It tries to flag aberrant behavior as a
  possible intrusion. The baseline profile can evolve over time, especially in
  wireless and mobile environments. Single events should therefore not result
  in absolute decisions.

  \item[Signature or misuse detection] starts with known patterns of attacks.
  It tries to find evidence of intrusions in the actions of and communication
  with a device. Because attacks often use valid ways to communicate with
  services on the devices, this approach can produce false positives.

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although this approach can detect previously
  unknown attacks, it can sometimes not detect specific attacks that anomaly or
  misuse detection can.

\subsection{Cooperative Decision Making}
\label{coorperative}

These taxonomies focus on the act of detecting intrusions themselves. In a
distributed environment, communication between participants may be even more
important.

Because not all devices are always actively participating in the network, they
can miss events that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important class of ID algorithms consists of cooperative
algorithms that combine information from single identities into a group-based
decision about alleged intrusions. We identify two ways to construct
cooperative decision making and focus on the way participants exchange
information: broadcast and interactive communication.

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally: devices can exchange information about the
  reputation of other devices, solely using one-way broadcast messages.
  Combining this information allows them to decide about their trust in a given
  device, based on more than their own, often partial, observations. The
  reputation-based detection algorithm introduced in
  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
  of a watchdog found in \cite{mishra2004intrusion}.

  \item[Interactive communication] allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of five
  phases, which essentially implement a voting system based on the Guy Fawkes
  protocol \cite{anderson1998new}, to identify an intruder. It enables the
  exchange of suspected intruder information and allows for distributed
  authentication of those \emph{votes}. Based on these authenticated votes, the
  participants can make a distributed decision about the jointly identified
  intruder.

\subsection{Properties and Scope of \NAME}
\label{scope}

Figure \ref{fig:classification} visually summarizes the dual classification and
scope of \NAME.

\begin{figure}[t]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  sibling distance=20mm,node distance=40mm
  ]
  \node (detection)             [align=center] {Detecting\\Intrusions}
    child {node (anomaly)       [align=center] {Anomaly}}
    child {node (signature)     [align=center] {Signature\\Misuse}}
    child {node (specification) [align=center] {Specification}};
  \node (decision)              [align=center,right=of detection] {Decision\\Making}
    child {node (broadcasting)  [align=center] {Broadcast}}
    child {node (interactive)   [align=center] {Interactive}};
\end{tikzpicture}
}
  \caption{Classification of ID Algorithms}
  \label{fig:classification}
\end{figure}

Algorithms that fit this classification are typically platform independent.
They look for patterns in network messages that can reveal intrusions and, in
doing so, all deal with the same information: other devices in the network. To
be able to formally describe ID algorithms and target the heterogeneous IoT,
\NAME requires ID algorithms to expose these properties.

A class of ID algorithms that does not have these properties is software
attestation \cite{seshadri2008sake}. It enables exchanging information about
the software that is running on a device and identifying devices with modified
content. These algorithms are highly platform dependent and have no data in
common with other algorithms, as they look for the effects of an intrusion, not
the intruder itself.

\section{A Design Pattern for ID Algorithms}
\label{pattern}

Based upon the classification presented in section \ref{classification}, we now
introduce a design pattern that identifies common functional aspects of ID
algorithms. It will allow us to define a strategy for optimizing the
combination of these algorithms. We extract two activities common to most
algorithms: the inspection of network traffic and the evaluation of thresholds.

Network traffic is the carrier along which malicious payloads reach a device
directly. Indirectly a device can also detect aberrant patterns of
communication by other devices, as a result of a successful attack.

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can do that in some cases, but anomaly
detection typically deals with some level of uncertainty. Also, in distributed
setups it is impossible to fully trust any neighboring device. In these
situations one has to rely on statistics and thresholds to decide when to drop
trust in a given device.

Updating such statistics and evaluating the corresponding thresholds should be
distinct operations. The example of a watchdog \cite{mishra2004intrusion}
illustrates this: a device monitors the actions of other devices and checks if
they correctly forward messages to distant nodes in the network. Missing
forwards can indicate a malicious intension to silently drop messages and
obstruct the normal operation of the network. There are valid reasons why a
device can miss the theoretical window in which one expects activity, i.e.\ the
wireless network being the most predominantly unreliable component. Directly
reacting upon a basic threshold can result in false positives. A way to
overcome this is to decouple the detection of problems and the decision what to
do about them.

This decoupling of detection and decision is the core of the design pattern for
ID algorithms. Listing \ref{alg:id-algo-pattern} presents the design pattern in
pseudo-code.

\begin{algorithm}[t]
  \caption{ID Algorithm Pattern}
  \label{alg:id-algo-pattern}
  \begin{algorithmic}
    \Require{devices: global variable for device information}
    \Function{process\_message}{$msg$}
      \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
       \State \dots \Comment{analyze byte-sequences}
      \EndFor
      \Let{$devices_x$}{value}  \Comment{optionally update device info}
      \State \Call{send}{$devices_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \State
    \Function{do\_housekeeping}{}
      \ForEach{$device \in devices$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$device > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device$, ``info''} \Comment{communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

Applying the pattern in the context of multiple algorithms, extends its scope
and a pattern such as described in listing \ref{alg:id-algo-application} will
typically appear.

\begin{algorithm}[t]
  \caption{Application pattern of multiple ID algorithms}
  \label{alg:id-algo-application}
  \begin{algorithmic}
    \Let{$msg$}{$network$} \Comment{all observed messages}
    \ForEach{$algorithm \in algorithms$}
      \State \Call{algorithm.progress\_message}{msg}
    \EndFor
    \State \dots
    \Comment{at a given interval}
    \ForEach{$algorithm \in algorithms$}
      \State \Call{algorithm.do\_housekeeping}{}
    \EndFor
  \end{algorithmic}
\end{algorithm}

The pattern calls the individual, linked implementations in sequence in two
situations: first, when the device observes new messages on the network and,
secondly, at a regular interval to allow for internal housekeeping, such as
threshold evaluation.

\begin{figure}[b]
  \centering
  \scalebox{1}{
\begin{tikzpicture}[
  node distance=9mm,
  ,>=stealth,very thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)   [user]                                      {DSL};
  \node (cg)    [representation,right=of dsl,align=center]  {Code\\Generator};
  \node (code)  [generated,right=of cg]                     {code};
  \node (comp)  [representation,right=of code]              {Compiler};
  \node (image) [generated,right=of comp]                   {image};

  \node (nodes) [dom,above=of cg, align=center]             {Nodes\\Domain};
  \node (plat)  [platform,below=of cg,align=center]         {Platform \&\\Language};          
  \node (lib)   [native,below=of comp,align=center]         {Software\\Library};

  \draw [->] (dsl)   -> (cg);
  \draw [->] (nodes) -> (cg);
  \draw [->] (plat)  -> (cg);
  \draw [->] (cg)    -> (code);
  \draw [->] (code)  -> (comp);
  \draw [->] (lib)   -> (comp);
  \draw [->] (comp)  -> (image);

\end{tikzpicture}
}
\caption{The high level design of \NAME, showing the resources, processes and
artifacts. A grayscale background represents the level of platform dependency,
with darker shades indicating a higher dependency. Both the source code and the
installable image are artifacts.}

\label{fig:design}
\end{figure}

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms,
but conflict with the target environment, the IoT\@. \NAME provides a framework
to efficiently implement an IDS for the IoT and offers a solution for the
identified problems. It consists of a 1) DSL, 2) source code generator and 3)
software library. Figure \ref{fig:design} visualizes the overall design.

% figure design is moved up to end up on this page

The DSL enables researchers to formally describe ID algorithms, without
focusing on any specific language or platform. A code generator can avoid the
identified problems concerning loops, common data and network access, as the
formal description provides the underlying intent, not the technical
implementation. It can also target different languages and platforms, while
reusing a software library that provides services including parsing, network
access wrapping,\dots

Using a DSL and code generator defines a development strategy. This has
implications, both on what is possible and how it can be achieved. We will
first position \NAME and motivate its design with respect to these trade offs.
Second, we will look at the principal design components of the DSL. Third, we
discuss the structure of the code generator and introduce its software library,
that offers a way to deal with for example the network access issue.

\subsection{\NAME as a Development Strategy}
\label{positioning}

The introduction of a code generator in the development process changes the
traditional way of working. This is most importantly due to the fact that the
actual source code is no longer the driving force that can be controlled by
developers. It is literally an intermediate representation between code
generator and compiler.

A code generator constructs this representation using three components, as
shown in figure \ref{fig:design}: 1) the description of the algorithm in the
DSL, 2) information about the domain, in this case the Nodes domain, and 3)
information about the platform and target language. At the level of the
compiler, a software library is introduced to offload specialized
implementations and allow custom extensions.

The applied grayscale indicates the components' platform dependency: the DSL is
platform and language independent and deals with subsystems, such as the
network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
Platform and language descriptions convert the abstract constructions in
actually supported code. The software library is standard native code,
targeting the platform and language. Finally, the code and image are artifacts
of the automated generation and compilation process. They aren't a product of
manual work and therefore aren't directly dependent on the platform.

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform dictates the level of effort to
implement changes in that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

\subsection{\NAME as a Domain Specific Language}
\label{dsl-design}

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \NAME's DSL and model focus on avoiding constructions
that result in problems concerning loops, common data and the use of the
network. To do this, \NAME uses an event-based approach to defining
functionality.

Besides avoiding these problems, \NAME's DSL also provides a higher level of
abstraction with respect to memory management and data access.

This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
3) unified messaging, 4) language support, and illustrates these with a
complete example.

\subsubsection{Loops}

Loops are a major contributor to the problem of algorithm fusibility. Loops in
program code are technical constructs and in fact often hide the intended
functionality.

The example of parsing network messages illustrates this. Each algorithm deals
with parsing in the same way: loop over the bytes in the payload and look for a
pattern that identifies what the algorithm is looking for. This is in fact a
technical implementation choice. The underlying functional intent is to react
when such a pattern is present in a message. The loop has \emph{activated} this
functionality, but the real goal is to \emph{passively} react on an event, not
\emph{actively} look for it. Event handling therefore can transform
constructions with loops to their functional meaning, allowing identification
and fusion of common functionality.

The same goes for actively polling some variable until it reaches a certain
threshold. When transformed to a reaction to an event, it eliminates the need
for a technical loop-construction.

\subsubsection{The Nodes Domain}

Most, if not all, ID algorithms deal with information about other devices in
the network. To share this information, \NAME provides a unified \ttt{nodes}
concept as part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This allows for only one
instance per known device to be present in memory. Finally, nodes can define a
scope for a function to execute in. This way, nodes offer a way to i.e.\
iterate all known devices, again without the need for an explicit loop.

\subsubsection{Unified Messaging}
\label{dsl-unified-msg}

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them can lower this overhead.

Sending and receiving of network messages is tightly integrated with the
previously mentioned concepts of event handling and the Nodes domain. An
underlying software library can handle all interaction with the network,
triggering functions by means of events when parsing new messages and exposing
an API to send messages through the Nodes domain. This approach gives the
framework complete control over communication and for example allows for
optimized marshaling and grouping of messages.

\subsubsection{Language Support}
\label{language-support}

Using a platform independent language offers the possibility to relieve the
implementer of explicit memory management and low-level data structures. From a
language perspective, \NAME offers several programming constructs, typically
found in modern programming languages with higher levels of abstraction:
optional typing, object orientation, pattern matching, aspect orientation.

\subsubsection{An Example}

>>>>

\title{
\FOO: A framework for efficient source code generation\\
of detection algorithms for the Internet of Things
}

\author[]{Christophe Van Ginneken}
\author[]{Jef Maerien}
\author[]{Christophe Huygens}
\author[]{\\Wouter Joosen}
\author[]{Danny Hughes}

\address{
iMinds-DistriNet, KU Leuven, 3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be
}

@@

The ability to detect patterns in its usage is what makes \emph{things}
\emph{smart}. The more patterns it can detect, the more accurate its
operational results will be. However, supporting an adequate set of pattern
detection algorithms imposes significant overhead on resource-constrained
devices such as those in the Internet of Things. Manual fusion of the
algorithms reduces resource consumption by optimizing resource sharing, yet,
proves to be time-consuming, repetitive and error-prone. To address this
problem we propose \FOO, a framework for the generation of efficient detection
algorithms. It consists of a domain specific language and a code generator. The
language allows formally describing the intent of detection algorithms and
supports the generator in organizing the source code. As a case study, we apply
\FOO to the generation of an Intrusion Detection System for wireless sensor
nodes. A side-by-side comparison shows that \FOO-generated source code reduces
message passing overhead, execution time and memory footprint in comparison to
sequential calls into stand-alone implementations of the algorithms.

@@

IoT, Domain Specific Languages, Source Code Generation, Source Code
Optimization, Intrusion Detection, Context Awareness

@@

The Internet of Things (IoT), the mother of all ubiquitous and pervasive
networks, populates our daily lives with millions of internet-connected
\emph{smart things}: from smart watches that assist visually impaired
persons\cite{porzi2013smart} to monitoring systems for flood hazardous
areas\cite{hughes2006gridstix}. Their applications cannot be more diverse,
however they share one common concern: each of these \emph{things} is
resource-constrained, relying on a battery and operating with limited
processing power and memory.

Although their applications at first seem unrelated, under the hood they share
a common paradigm: pattern recognition. With the introduction of \emph{smart
things}, we want to augment our lives with information that is hard, or at
least time and resource consuming, to obtain otherwise. The ability to detect
patterns in its usage or its applied context, is what makes these \emph{things}
\emph{smart}. A watch that recognizes gestures can perform certain actions on
behalf of its wearer. That same watch can also detect a wet floor and notify
this hazard.

Monitoring human activities is an active research field for more than a
decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
the real world. Equipped with wearables with numerous sensors, the
possibilities seem endless, but so are the number of detection and recognition
algorithms that have to run simultaneously on these small \emph{things}, to
support our augmented reality.

Another prime example of a class of algorithms that inhibit the same properties
related to pattern recognition is Intrusion Detection (ID). The vast number of
security related opportunities for attackers makes way for an equally sized
amount of ID algorithms, all with a similar structure, but largely independent
operation.

@@

The contributions of this paper are threefold: The first contribution consists
in the formulation of a design pattern for detection/recognition algorithms for
IoT devices. This defines the target algorithms that fit our proposed
framework. The second contribution is a framework called \FOO.\@ It enables the
formal description of algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \FOO also addresses the heterogeneous nature of both IoT and the
algorithms. The third contribution introduces functional code fusion (FCF) as a
source code generation paradigm to address the combination of algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
identifies problems that arise when combining multiple algorithms in the
context of resource-constrained devices. Section \ref{pattern} formulates a
pattern common to the detection/recognition algorithms in the IoT context.
Section \ref{design} introduces our framework, its design and components. In
section \ref{evaluation} we apply our solution to case study related to ID,
showing an actual implementation and evaluation of the experimental results.
Section \ref{related} explores related work in the field of DSLs and code
generation. Finally, section \ref{conclusion} summarizes our findings, draws
conclusions and identifies topics for future work.

\section{Problems when combining multiple algorithms}
\label{problems}

Simply taking existing implementations of algorithms and calling them in
sequence, to allow them to perform their functionality, will in most cases
result in suboptimal code. Three prime examples on the source code level
illustrate this:

\vspace{-2mm}

@@

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can cause overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns. These loops can also hold calls to poll sensor for
  their values. This results in an execution overhead, putting a strain on the
  power consumption.

\vspace{-2mm}

  \item[Common data] cross-cuts the different algorithms, that have to store
  the same information over and over again within their own scope. This results
  in a lot of duplicated information in memory, raising the hardware
  requirements.

\vspace{-2mm}

  \item[The network] allows algorithms to operate on a distributed scale.
  Scattered calls to a message sending function cause multiple accesses to the
  wireless radio, keeping it active and energy consuming more than strictly
  needed.

@@

Two operational aspects are orthogonal to these problems and further amplify
them:

\vspace{-2mm}

@@

  \item[Reimplementing] algorithms manually to eliminate these issues is
  not a one-time cost. A release of a new version of an algorithm, or simply a
  reselection of algorithms, due to a changing requirements, can trigger it.
  Also, changing, or simply implementing the algorithms from scratch, holds the
  risk of making mistakes.

\vspace{-2mm}

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This partially explains why
  there are few implementations of algorithms readily available to developers:
  there is no single platform to target for researchers, which lowers the
  opportunity to reuse prior work.

@@

\section{A Design Pattern for Detection/Recognition Algorithms}
\label{pattern}

The source code level problems that arise when combining multiple algorithms,
as presented in section \ref{problems}, are in fact inherent to the design
pattern that is the foundation for the type of algorithms we identify in
typical detection/recognition applications in IoT applications. The formulation
of such a design pattern will allow us to define a strategy for the
optimization of the combination of these algorithms. We start by extracting two
common activities of these algorithms: the collection of sensor values and the
evaluation of thresholds. Periodically collecting sensor values requires
preparing the hardware for value extraction, waiting for the value to become
available, reading the actual bytes, interpreting them into a meaningful value
and recording them for further analysis. This can be as simple as reading a
single byte, but can also be as complex as parsing a network message.

A single recorded sensor value hardly ever results in a decision. The
combination, often over time, of several of these values and the evaluation of
thresholds will be key to decide if a change in the environment requires an
action. Updating such statistics and evaluating the corresponding thresholds
should be distinct operations. Therefore this is a separate function that is
also periodically executed.

This decoupling of detection and decision is the core of the design pattern for
ID algorithms. Listing \ref{alg:pattern} presents the design pattern in
pseudo-code. Applying the pattern in the context of multiple algorithms,
extends its scope and a pattern such as described in listing
\ref{alg:application} will typically appear. The pattern calls the individual,
implementations in sequence in two situations: first, to collect sensor
readings, and secondly, at a regular interval to allow for internal
housekeeping, such as threshold evaluation.

We clearly see the three aforementioned problems manifest itself: first, the
nesting of loops causes redundant executions of potentially the same sensor
readings inside the individual algorithms' $collect\_data$ functions. Second,
each individual algorithm stores the same sensor values in memory. Third,
scattered calls to e.g. network functions (in case of distributed algorithms)
lead to suboptimal use of the wireless radio.

Implementing the solution from scratch, it would be possible to identify these
problems and reorganize the code in such a way that sensor values are only read
once, stored in memory once and network messages parsed once. In section
\ref{problems} we identified two operational problems that show that this is
not feasible or at least suboptimal. We therefore present \FOO, a framework
that enables formally describing such algorithms and fusing common data and
functions in an automated way, to overcome both these problems.

\vspace{-2mm}

\begin{algorithm}[b]
  \caption{Detection/Recognition Algorithm Pattern}
  \label{alg:pattern}
  \begin{algorithmic}
    \Require{readings: global variables to hold collected sensor data for the given algorithm}
    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
    \Function{collect\_data}{}
      \ForEach{$sensor \in required\_sensors$}
        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
      \EndFor
      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \Function{do\_housekeeping}{}
      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$reading > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[b]
  \caption{Application pattern of multiple algorithms}
  \label{alg:application}
  \begin{algorithmic}
    \ForEach{$algorithm \in algorithms$}
      \State \Call{algorithm.collect\_data}{}
    \EndFor
    \ForEach{$algorithm \in algorithms$}        \Comment{at a given interval}
      \State \Call{algorithm.do\_housekeeping}{}
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to detection
algorithms, but conflict with the target environment, the IoT\@. \FOO provides
a framework to efficiently implement these algorithms for the IoT and offers a
solution for all identified problems. It consists of a 1) DSL, 2) source code
generator and 3) software library. Figure \ref{fig:design} visualizes the
design.

\begin{figure}[b]
  \centering
  \scalebox{0.75}{
\begin{tikzpicture}[
  node distance=9mm,
  ,>=stealth,very thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)     [user]                                      {DSL};
  \node (cg)      [representation,right=of dsl,align=center]  {Code\\Generator};
  \node (code)    [generated,right=of cg]                     {code};
  \node (comp)    [representation,right=of code]              {Compiler};
  \node (image)   [generated,right=of comp]                   {image};

  \node (domains) [dom,above=of cg, align=center]             {Functional\\Domains};
  \node (plat)    [platform,below=of cg,align=center]         {Platform \&\\Language};          
  \node (lib)     [native,below=of comp,align=center]         {Software\\Library};

  \draw [->] (dsl)     -> (cg);
  \draw [->] (domains) -> (cg);
  \draw [->] (plat)    -> (cg);
  \draw [->] (cg)      -> (code);
  \draw [->] (code)    -> (comp);
  \draw [->] (lib)     -> (comp);
  \draw [->] (comp)    -> (image);
\end{tikzpicture}
}
\caption{The high level design of \FOO, showing the resources, processes and
artifacts. A grayscale background represents the level of platform dependency,
with darker shades indicating a higher dependency. Both the source code and the
installable image are artifacts.}
\label{fig:design}
\end{figure}

The DSL enables researchers to formally describe algorithms, without focusing
on any specific language or platform. A code generator can avoid the identified
problems concerning loops, common data and network access, as the formal
description provides the underlying intent, not the technical implementation.
It can also target different languages and platforms, while reusing a software
library that provides services including parsing, network access wrapping,\dots
It applies the Functional Code Fusion (FCF) paradigm to identify the hotspots
for optimization. A key component of FCF is the introduction of
\emph{functional domains} that harbor common functionality.

Using a DSL and code generator defines a development strategy. We will first
position \FOO and motivate its design. Second, we will look at the principal
design components of the DSL.\@ Third, we discuss the structure of the code
generator and introduce its software library, that offers a way to deal with
for example the network access issue.

\vspace{-1mm}
\subsection{\FOO as a Development Strategy}
\label{positioning}

The introduction of a code generator in the development process changes the way
of working. The actual source code is no longer the driving force, controlled
by developers. It is literally an intermediate representation between code
generator and compiler. A code generator constructs this representation using
three components, as shown in figure \ref{fig:design}: 1) the description of
the algorithm in the DSL, 2) information about the functional domains, and 3)
information about the platform and target language. At the level of the
compiler, a software library offloads specialized implementations and allow
custom extensions. The applied grayscale indicates the components' platform
dependency. This hierarchy is important from a development effort point of
view. The level of dependency on the target platform dictates the level of
effort to implement changes. This is important in relation to the scalability
of the framework itself.

\vspace{-1mm}
\subsection{\FOO as a Domain Specific Language}
\label{dsl-design}

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \FOO's DSL and model focus on avoiding constructions that
result in problems concerning loops, common data and the use of the network. To
do this, \FOO uses an event-based approach to defining functionality.
Describing algorithms in function of events that occur will remove the
technical descriptions and allows FCF to identify the actual intent of the
algorithms with respect to the identified source code problems.

\vspace{-1mm}
\subsubsection{Loops}

Loops are a major contributor to the problem of algorithm fusibility. Loops in
program code are technical constructs and in fact often hide the intended
functionality. The example of parsing network messages illustrates this. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional intent is to react when such a pattern is present in a message. The
loop has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it. Event
handling therefore can transform constructions with loops to their functional
meaning, allowing identification and fusion of common functionality.

\vspace{-1mm}
\subsubsection{Common Data}

Different detection algorithms typically deal with the same classes of
information, be it sensor data or network messages. Functional domains
consolidate these and store and manage this data in a unified way. \FOO
provides the concept of such domains as plug-ins to the DSL, extending it with
domain specific storage and logic. For example in the case of distributed
algorithms, such as ID algorithms, a \emph{Nodes} domain will manage
information about neighbor nodes, exposing them as global objects to the
algorithms, relieving them from managing these themselves. Functional domains
are accessible as collections, allowing execution of functions in their scope,
eliminating the need for explicit loops.

\vspace{-1mm}
\subsubsection{Unified Messaging}
\label{dsl-unified-msg}

Access to the network, scattered around an algorithm, causes the wireless radio
to be more active than needed. Delaying the sending of messages and grouping
them can lower this overhead. Sending and receiving of network messages is
tightly integrated with the previously mentioned concepts of event handling and
for example the Nodes domain. An underlying software library can handle all
interaction with the network, triggering functions by means of events when
parsing new messages and exposing an API to send messages through the Nodes
domain. This approach gives the framework complete control over communication
and allows for optimized marshaling and grouping of messages.
